Type-Safe Modular Parsing

### 2017-05-15 SLE_17/sections/Discussion.tex
More.

Using Object Algebras, it is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters. As an example, we extend the language of literals, additions and variables by introducing types, including primitive type \inlinecode{int} and function (arrow) type, and we add lambda abstractions to expressions. Now the language has two sorts in the syntax, types and expressions:

\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>

@@

Figure~\ref{fig:multi} illustrates the corresponding Scala code
that extends the Object Algebra interface, pretty-printing operation and parser. We use two type parameters \inlinecode{E} and \inlinecode{T} for expressions and types. They guarantee that invalid terms such as \inlinecode{int + int} will be rejected by the parser.
Besides lexing, the trait \inlinecode{LamOAParser} also introduces parsers for types, and the new case for expressions.
We use \inlinecode{pTypedLamT} and \inlinecode{pTypedLamE} as copies of current \inlinecode{pT} and \inlinecode{pE}, due to the issue
with \inlinecode{super} in Scala (see discussion in Section~\ref{subsec:parsingwithoa}). \inlinecode{pT} and \inlinecode{pE} are the parsers used for performing recursion.

\begin{figure}[ht]
\lstinputlisting[linerange=7-33]{code/src/papercode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

From the code, we observe that the multi-sorts of Object Algebra interface
express multiple syntax. The isolation for different syntax
is guaranteed by the type system. The following client code

\lstinputlisting[linerange=40-43]{code/src/papercode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
works as expected to parse and pretty-print the result.

@@

As many syntactically extensible parsers, our approach also supports
modifying part of existing parsers, including updating or eliminating existing rules,
but in a type-safe way. This can be useful in many
situations, for instance when conflicts or ambiguities arise upon composing languages.
As an illustration, suppose we have an untyped lambda abstraction case in a base parser as below. Some irrelevant details are omitted.

\lstinputlisting[linerange=76-82]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=BASEPARSER_UNTYPEDLAM
Here \inlinecode{pLam} parses a lambda symbol, an identifier, a dot and an expression in sequence.
Then we want to replace the untyped lambda abstractions by typed
lambdas. With inheritance and method overriding, it is easy to only
change the implementation of \lstinline{pLam} in the extended parser.
Due to dynamic dispatch, our new
implementation of lambdas will be different without affecting the other parts of the parser.

\lstinputlisting[linerange=86-92]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=EXTPARSER_TYPEDLAM

@@

We take the example in Section~\ref{subsec:parsingwithoa} again, which is a language of literals, additions and variables. It can be defined as a language component \inlinecode{VarExpr}.

\lstinputlisting[linerange=115-130]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR

@@

\lstinputlisting[linerange=134-150]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM

@@

\lstinputlisting[linerange=154-165]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR

@@

Under our modular parsing framework, we use Packrat parsing as the underlying parsing technique, OO inheritance with method overriding for composing and extending parsers, and Object Algebras for parsing extensible ASTs. However, such a pattern is more general and thus more powerful, because those aspects are orthogonal to each other, and hence can have alternatives.

\begin{itemize}[leftmargin=*]

\item {\bf Parsing Technique:}
The pattern does not depend on a particular parser combinator library
or parsing algorithm. We demonstrated that Packrat parsing has some
advantages and is suitable in a modular setting. However, under
other circumstances it may not be
the best choice. One can use other parser combinators as long
as the guidelines proposed in Section~\ref{sec:packrat} are met.

\item {\bf Composing and Extending Parsers:}
  We use traits to model parsers, OO inheritance to compose them, and
  method overriding for their extensibility. These features
  may not be supported in other programming languages, especially in
  functional languages such as Haskell. Nevertheless, \textit{open
    recursion}~\cite{CookThesis} could be used as an alternative. An explicit
  ``self-reference'' parameter is able to explain the recursive calls
  dynamically, by the real argument passed at runtime.

\item {\bf Extensible ASTs:}
Besides Object Algebras, many other techniques including \textit{Data
  types à la carte} (DTC)~\cite{swierstra2008data} and the Cake pattern~\cite{odersky2005independently} also support extensible data structures. These alternatives could also be adopted to build ASTs and their parsers.

\end{itemize}

As we demonstrated, such a pattern for modular parsing is 
modular itself, and can be customized easily.

>>>>

Using Object Algebras, it is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters.
For instance, now we extend the expression language from the end of Section~\ref{sec:algebrasandparsing}, with a
primitive type \inlinecode{int} type and typed lambda abstractions:

\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int'

@@

Figure~\ref{fig:multi} illustrates the corresponding Scala code
that extends the Object Algebra interface, pretty-printing operation and parser. We use two type parameters \inlinecode{E} and \inlinecode{T} for expressions and types. The type system guarantees that invalid terms such as \inlinecode{int + int} will be rejected.
Besides lexing, the trait \inlinecode{LamOAParser} also introduces parsers for types, and the new case for expressions.
We use \inlinecode{pTypedLamT} and \inlinecode{pTypedLamE} as copies of current \inlinecode{pT} and \inlinecode{pE}, due to the issue
with \inlinecode{super} in Scala (see discussion in Section~\ref{subsec:parsingwithoa}). \inlinecode{pT} and \inlinecode{pE} are the parsers used for performing recursion.

\begin{figure}[ht]
\lstinputlisting[linerange=7-23]{code/src/papercode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

@@

As many syntactically extensible parsers, our approach also supports
modifying part of existing parsers, including updating or eliminating existing rules,
but in a type-safe way. This can be useful in many
situations, for instance when conflicts or ambiguities arise upon composing languages.
As an illustration, suppose we have an untyped lambda abstraction case in a base parser, defined as a value:
\begin{lstlisting}
val pLam: Parser[E] =
  ("\\" ~> ident) ~ ("." ~> pE) ^^ ...
\end{lstlisting}
Here \inlinecode{pLam} parses a lambda symbol, an identifier, a dot and an expression in sequence.
Then we want to replace the untyped lambda abstractions by typed
lambdas. With inheritance and method overriding, it is easy to only
change the implementation of \lstinline{pLam} in the extended parser.
Due to dynamic dispatch, our new
implementation of lambdas will be different without affecting the other parts of the parser.
\begin{lstlisting}
override val pLam: Parser[E] =
  ("\\" ~> ident) ~ (":" ~> pT) ~ ("." ~> pE) ^^ ...
\end{lstlisting}

@@

We take the example in Section~\ref{subsec:parsingwithoa} again. It can be defined as a language component \inlinecode{VarExpr}.
For space reasons we omit some detailed code.

\lstinputlisting[linerange=115-124]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR

@@

\lstinputlisting[linerange=128-137]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM

@@

\lstinputlisting[linerange=141-152]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR

@@

Our prototype uses Packrat parsing as the underlying parsing technique, OO inheritance for composing and extending parsers, and Object Algebras for parsing extensible ASTs. Yet such a framework is itself flexible and modular, because those techniques can have alternatives.
For example, as we mentioned before, any parsing library
that resolves the algorithmic challenges in modular parsing can work well. Regarding OO inheritance for the extensibility, an alternative approach, called \textit{open recursion}~\cite{CookThesis} can be used in other languages, by introducing explicit ``self-reference'' parameters for the recursion. Furthermore, besides Object Algebras, v\textit{Data types à la carte} (DTC)~\cite{swierstra2008data} and the Cake pattern~\cite{odersky2005independently} also support extensible data structures. For the goal of modular parsing a custom combination of those alternatives can be adopted.

### 2017-05-15 SLE_17/sections/Inheritance.tex
More.

\subsection{The Mismatch Between OO ASTs and Grammars}
Object-oriented ASTs are ASTs encoded via standard OO techniques,
such as the {\sc Composite} or {\sc Interpreter} patterns~\cite{gamma1995design}. Such ASTs
fix the operations required in the AST, but allow the easy addition of new
language constructs. This is in contrast with ASTs modelled using
functional programming algebraic datatypes or OO style Visitors, where
adding new operations is easy, but adding new language constructs is hard.
In Section~\ref{subsec:packratparsing} we provide a simple example of OO ASTs. The same abstract syntax could be encoded using algebraic datatypes (or alternatively Visitors)
as:

@@

Algebraic datatypes and grammars fit well together. Indeed there are
obvious similarities in declaring a grammar, and declaring an
algebraic datatype. It is difficult to modularly add new language
constructs to both grammars and ASTs based on algebraic datatypes. However,
ASTs modelled with algebraic datatypes naturally allow for the easy
addition of operations. So using such ASTs at least preserves one
dimension of extensibility.

@@

\lstinputlisting[linerange=46-50]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

The \inlinecode{super[T].x} syntax in Scala, so called \textit{static super reference}, refers to the type or method \inlinecode{x} in the parent trait \inlinecode{T}. Under multiple inheritance, it can be used to distinguish the methods of the same name. Therefore in the new parser, we use \inlinecode{super} to specify which \inlinecode{pExpr} we are referring to.

As a concrete example, we compose the language of integers and additions with a language of boolean values and if-then-else expressions, to build a new language.

\lstinputlisting[linerange=56-73]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=BOOL_EXAMPLE

@@

Then the new parser, which contains both integer and boolean
expressions, is able to parse hybrid expressions such as
\inlinecode{"if true then 1 + 2 else 3 + 4"}.

>>>>

\subsection{The Mismatch Between OO ASTs and Grammars}\label{subsec:mismatch}
Object-oriented ASTs are ASTs encoded via standard OO techniques,
such as the {\sc Composite} or {\sc Interpreter} patterns~\cite{gamma1995design}. Such ASTs
fix the operations required in the AST, but allow the easy addition of new
language constructs. We have seen an example of OO ASTs in Section~\ref{subsec:packratparsing}.
In contrast, ASTs modelled using functional programming algebraic datatypes (or alternatively OO style Visitors) are as follows:

@@

Algebraic datatypes and grammars fit well together. Indeed there are
obvious similarities in declaring a grammar, and declaring an
algebraic datatype. Unlike OO ASTs, they naturally allow for the easy
addition of operations, whereas modularly adding new constructs to grammars
or ASTs becomes hard. Using this style of ASTs again preserves one dimension of extensibility.

@@

\lstinputlisting[linerange=46-50]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE
The \inlinecode{super[T].x} syntax in Scala, so called \textit{static super reference}, refers to the type or method \inlinecode{x} in the parent trait \inlinecode{T}. Under multiple inheritance, it can be used to distinguish the methods of the same name. Therefore in the new parser, we use \inlinecode{super} to specify which \inlinecode{pExpr} we are referring to.

%As a concrete example, we compose the language of integers and additions with a language of boolean values and if-then-else expressions, to build a new language.

%\lstinputlisting[linerange=56-68]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=BOOL_EXAMPLE

@@

%Then the new parser, which contains both integer and boolean
%expressions, is able to parse hybrid expressions such as
%\inlinecode{"if true then 1 + 2 else 3 + 4"}.

### 2017-05-15 SLE_17/sections/ObjectAlgebras.tex
More.

The Expression Problem~\cite{wadler1998expression} illustrates the
difficulty of extending data structures or ASTs in two dimensions. For
traditional OO ASTs, adding new language constructs is easy, but
adding new operations is hard. In the last section, we have a simple
language containing literals, additions and variables. The
corresponding ASTs are built using classes with a pretty-printing
operation.
To modularly add an operation like collecting free variables, one attempt
would be extending
\inlinecode{Expr} with the new operation to obtain a new abstract type
for ASTs:

@@

\noindent Then all classes representing language constructs could be
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
even if that approach would work, the parsing code would not
be reusable. \lstinline{VarParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser have type \inlinecode{Parser[Expr]},
which generates ASTs that only support pretty-printing. Now we have to substitute \inlinecode{NewExpr} for \inlinecode{Expr} (the
same for \inlinecode{Lit}, \inlinecode{Add}, and so on). It requires either modification on existing code or casts.
The goal of semantic modularity motivates us to find a different approach for building ASTs.

Fortunately, Object Algebras~\cite{Oliveira:2012} enable us to solve
this problem. Object Algebras separate data variants and
operations, and offer high flexibility in the choice of operations to
be performed over ASTs. The next section provides a brief introduction
to Object Algebras.

@@

Object Algebras capture a design pattern that addresses the Expression
Problem, achieving two dimensions of extensibility (language
constructs and operations) in a modular and type-safe way. The
definition of data structures is separated from their behaviours, and
future extensions on both dimensions no longer require existing code
to be modified, supporting separate compilation.

@@

\lstinputlisting[linerange=12-16]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression language:

\lstinputlisting[linerange=20-23]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EVAL

@@

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty-print the expression, the code

\begin{lstlisting}
makeExp(new VarPrint {})
\end{lstlisting}
results in \inlinecode{"(1 + x)"} as expected.

@@

\paragraph{Extensions}
Now let's try this pattern with the variables extension:

>>>>

The Expression Problem~\cite{wadler1998expression} illustrates the
difficulty of extending data structures or ASTs in two dimensions.
As we have mentioned in Section~\ref{subsec:mismatch}, the use of traditional ASTs
makes it hard to add new operations. In the last section we have seen a language
that supports pretty-printing (in \inlinecode{Expr}).
To modularly add an operation like collecting free variables, one attempt
would be extending
\inlinecode{Expr} with the new operation to obtain a new abstract type
for ASTs:

@@

\noindent Then all classes representing language constructs could be
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
even if that approach would work, the parsing code in \lstinline{VarParser} is no longer reusable!
The types \inlinecode{Expr, Lit, Add}, and so on, are all old types without the free variables operation.
To match the new ASTs, we have to substitute \inlinecode{NewExpr} for \inlinecode{Expr} (the
same for \inlinecode{Lit}, \inlinecode{Add}, ...). It requires either modification on existing code or casts.
The goal of semantic modularity motivates us to find a different approach for building ASTs.

%Object Algebras separate data variants and
%operations, and offer high flexibility in the choice of operations to
%be performed over ASTs. The next section provides a brief introduction
%to Object Algebras.

@@

Fortunately, Object Algebras~\cite{Oliveira:2012} enable us to solve
this problem.
They capture a design pattern that addresses the Expression
Problem, achieving two dimensions of extensibility (language
constructs and operations) in a modular and type-safe way. The
definition of data structures is separated from their behaviours, and
future extensions on both dimensions no longer require existing code
to be modified, supporting separate compilation.

@@

\lstinputlisting[linerange=12-16]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can also define an evaluation operation as a new trait that extends \inlinecode{Alg[Int]}. Hence adding new operations is modular. We omit that code
due to space reasons.

%\lstinputlisting[linerange=20-23]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EVAL

@@

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty-print the expression, the code \inlinecode{"makeExp(new VarPrint \{\})"}
results in \inlinecode{"(1 + x)"} as expected.

@@

\paragraph{Extensions}
Now let's try on the variables extension:

### 2017-05-15 SLE_17/sections/PackratParsing.tex
More.

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing.
One reason is that they are naturally modular for parser composition, providing full control to
the programmer. The other one is that they are implemented by a piece of code, so that it ensures
type safety in a statically typed programming language.
Unfortunately many parser combinators have important limitations.
In particular, several parser combinators including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

\paragraph{Left-Recursion Elimination} The top-down, recursive descent parsing strategy adopted by those parser combinator libraries cannot support left-recursive grammars directly. The common solution is to rewrite the grammar into an equivalent but non-left-recursive one, called left-recursion elimination.

The grammar below represents a simple arithmetic language containing only integers. Its corresponding parser is shown on the right side, written in Parsec.

\begin{minipage}{10em}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int>
\end{grammar}
\end{minipage}
\begin{minipage}{10em}
\begin{lstlisting}[language=PlainCode]
parseExpr = parseInt
\end{lstlisting}
\end{minipage}

Consider extending the language by adding subtractions to the grammar.
If we directly extend the grammar, the new left-recursive case will cause an infinite loop.
Below, \inlinecode{parseExpr} and \inlinecode{parseSub} call each other and never stop.

\begin{minipage}{12em}
\setlength{\grammarindent}{4em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `-' <int>
\end{grammar}
\end{minipage}
\begin{minipage}{15em}
\begin{lstlisting}[language=PlainCode]
parseExpr =
  parseSub <|> parseInt

parseSub = do
  e <- parseExpr
  ...
\end{lstlisting}
\end{minipage}

Instead, we have to rewrite the grammar as below to eliminate left-recursion.

@@

After left-recursion elimination, the structure of grammar is changed,
as well as its corresponding parser. In a modular setting, it is
possible but unnecessarily complicated to analyse the grammar and
rewrite it when doing extensions. Anticipating that every non-terminal
has left-recursive rules is helpful for extensibility but overkill,
since it is inconvenient and introduces extra complexity for
representation of grammars and implementation of parsers.

@@

In this case, reordering the alternatives ensures that
the \emph{longest match} is picked among the possible results. However, manual reordering for the longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers are supposed to \emph{manually} adjust
the order of parsers, which requires rewriting previously written code.

@@

Similarly, this violates a modular setting because we cannot have a global view of the full grammar.
Hence the worst case should be considered that all alternatives may share common prefixes with future cases. In that case we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere. However this results in the worst-case exponential time
complexity.

@@

\lstinputlisting[linerange=9-17]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST

@@

\lstinputlisting[linerange=21-28]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR

@@

\begin{table}
\begin{tabularx}{\linewidth}{X}
\hline
\lstinline{def ~[U](q: => Parser[U]): Parser[~[T, U]]} \\
- A parser combinator for sequential composition. \\ \hline
\lstinline{def \^\^[U](f: (T) => U): Parser[U]} \\
- A parser combinator for function application. \\ \hline
\lstinline{def \^\^\^[U](v: => U): Parser[U]} \\
- A parser combinator that changes a successful result into the specified value. \\ \hline
\lstinline{def <~[U](q: => Parser[U]): Parser[T]} \\
- A parser combinator for sequential composition which keeps only the left result. \\ \hline
\lstinline{def ~>[U](q: => Parser[U]): Parser[U]} \\
- A parser combinator for sequential composition which keeps only the right result. \\ \hline
\lstinline{def ident: Parser[String]} \\
- A parser which matches an identifier. \\ \hline
\lstinline{def numericLit: Parser[String]} \\
- A parser which matches a numeric literal. \\ \hline
\lstinline{def |[U >: T](q: => Parser[U]): Parser[U]} \\
- A parser combinator for alternative composition. \\ \hline
\lstinline{def |||[U >: T](q0: => Parser[U]): Parser[U]} \\
- A parser combinator for alternative with longest match composition. \\ \hline
\end{tabularx}
\caption{Common combinators from the Scala standard parser combinator library.}\label{tab:packrat}
\end{table}

>>>>

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
For the goal of modular parsing, parser combinators seem suitable because they are
naturally modular for parser composition, but also they ensure type safety.
Unfortunately many parser combinators have important limitations.
In particular, several parser combinators including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

\paragraph{Left-Recursion Elimination} The top-down, recursive descent parsing strategy adopted by those parser combinator libraries cannot support left-recursive grammars directly. For instance, we start with a simple arithmetic language containing only integers and subtractions. The
grammar with concrete syntax and part of the parsing code in Parsec are presented below:

\begin{minipage}{12em}
\setlength{\grammarindent}{4em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `-' <int>
\end{grammar}
\end{minipage}
\begin{minipage}{15em}
\begin{lstlisting}[language=PlainCode]
parseExpr =
  parseSub <|> parseInt
parseSub = do
  e <- parseExpr ...
\end{lstlisting}
\end{minipage}
Such a left-recursive implementation will cause an infinite loop, since \inlinecode{parseExpr} and \inlinecode{parseSub} call each other and never stop.  
A common solution is to rewrite the grammar into an equivalent but non-left-recursive one, called left-recursion elimination:

@@

\noindent After left-recursion elimination, the structure of grammar is changed,
as well as its corresponding parser. In a modular setting, it is
possible but unnecessarily complicated to analyse the grammar and
rewrite it when doing extensions. Anticipating that every non-terminal
has left-recursive rules is helpful for extensibility but overkill,
since it is inconvenient and introduces extra complexity for
representation of grammars and implementation of parsers.

@@

In this case, reordering the alternatives ensures that
the \emph{longest match} is picked among the possible results. However, manual reordering for the longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers are supposed to \emph{manually} adjust
the order of parsers, by rewriting previously written code.

@@

Similarly, this violates a modular setting because it also requires a global view of the full grammar.
Hence the worst case should be considered that all alternatives may share common prefixes with future cases. In that case we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere. However this results in the worst-case exponential time
complexity.

@@

\lstinputlisting[linerange=9-15]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST

@@

\lstinputlisting[linerange=19-26]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR

@@

\begin{table}
\begin{tabularx}{\linewidth}{X}
\hline
\lstinline{def ~[U](q: => Parser[U]): Parser[~[T, U]]} \\
- A parser combinator for sequential composition. \\ \hline
\lstinline{def \^\^[U](f: (T) => U): Parser[U]} \\
- A parser combinator for function application. \\ \hline
%\lstinline{def \^\^\^[U](v: => U): Parser[U]} \\
%- A parser combinator that changes a successful result into the specified value. \\ \hline
\lstinline{def <~[U](q: => Parser[U]): Parser[T]} \\
- A parser combinator for sequential composition which keeps only the left result. \\ \hline
\lstinline{def ~>[U](q: => Parser[U]): Parser[U]} \\
- A parser combinator for sequential composition which keeps only the right result. \\ \hline
\lstinline{def ident: Parser[String]} \\
- A parser which matches an identifier. \\ \hline
\lstinline{def numericLit: Parser[String]} \\
- A parser which matches a numeric literal. \\ \hline
\lstinline{def |[U >: T](q: => Parser[U]): Parser[U]} \\
- A parser combinator for alternative composition. \\ \hline
\lstinline{def |||[U >: T](q0: => Parser[U]): Parser[U]} \\
- A parser combinator for alternative with longest match composition. \\ \hline
\end{tabularx}
\caption{Common combinators from the Scala standard parser combinator library.}\label{tab:packrat}
\end{table}
